# Discrete-time Markov chains


Recall the simple random walk model of the price of the stock. Suppose
the stock price for the first four days are
$$(X_0, X_1, X_2, X_3) = (100, 99, 98, 99).$$ Based on this past
information, what can we say about the price at day 4, $X_4$? Although,
we completely know the whole past price history, the only information
relevant for predicting their future price is the price on the previous
day, i.e. $X_3$. This means that
$$\Pr(X_{4} = j | X_0 = 100, X_{1} = 99,  X_2 = 98 ,  X_3 = 99) = \Pr(X_{4} = j | X_3 = 99).$$
Given the current price $X_3$, the price $X_4$ at day 4 is independent
of the history prices $X_0, X_1, X_2$. The sequence of stock prices
$X_0, X_1, \ldots, X_n$ is an example of a **Markov chain**.

A Markov process is a special type of stochastic processes with the
property that the future evolution of the process depends only on its
current state and not on its past history. That is given the value of
$X_t$, the values of $X_s$ for $s > t$ do not depend on the values of
$X_u$ for $u < t$. This property is called the **Markov property**.

A **discrete-time Markov chain** is a discrete-time stochastic process
that satisfies the Markov property:
$$\Pr(X_{n+1} = j | X_n = i, X_{n-1} = i_{n-1}, \ldots, X_0 = i_{0}) = \Pr(X_{n+1} = j | X_n = i),$$
for all time points $n$ and all states $i_0, i_1, \ldots, i_{n-1},i,j$.
It is convenient to assume that the state space of the Markov chain is a
subset of non-negative integers, i.e. $S  \subseteq \{0, 1,  \ldots \}$.

::: {.example}
*A process with independent increments has the Markov
property.*
:::
**Solution:**

Recall the following definitions. An increment of a process is the
amount by which its value changes over a period of time, for e.g.
$X_{t +u} - X_t$ where $u > 0$.

A process $X_t$ is said to have independent increments if for all $t$
and every $u > 0$, the increment $X_{t +u} - X_t$ is independent of all
the past of the process $\{X_s : 0 \le s \le t \}$).

In order to show that a process with independent increments has the
Markov property, we proceed as follows: for all times $s_1 < s_2 < \ldots < s_n < s < t$ 
$$\begin{aligned}
\Pr(X_t \in A | X_{s_1} = x_1, X_{s_2} = x_2, \ldots, X_{s_n} = x_n ,X_{s} = x) 
&= \Pr(X_t - X_s + x \in A | X_{s_1} = x_1, X_{s_2} = x_2, \ldots, X_{s_n} = x_n, X_{s} = x) \\
&= \Pr(X_t - X_s + x \in A |  X_{s} = x)  \text{(by independence of the past)} \\
&= \Pr(X_t  \in A |  X_{s} = x).\end{aligned}$$

**Note** The random walk process has the Markov property.

## One-step transition probabilities

The conditional probability that $X_{n+1}$ is in state $j$ given that
$X_n$ is in state $i$ is called **one-step transition probability** and
is denoted by $$\Pr(X_{n+1} = j | X_n = i) = p_{ij}^{n,n+1}.$$ Note that
the transition probabilities depend not only on the current and future
states, **but also on the time of transition $n$**.

If the transition probabilities $p_{ij}^{n,n+1}$ in a Markov chain do
not depend on time $n$, the Markov chain is said to be
**time-homogeneous or stationary or simply homogeneous**. Then
$$p_{ij}^{n,n+1} = \Pr(X_{n+1} = j | X_n = i)  = \Pr(X_{1} = j | X_0 = i)  = p_{ij}.$$
Otherwise, it is said to be **nonstationary** or **nonhomogeneous**.

Unless stated otherwise, it shall be assumed that the Markov chain is
stationary. The matrix $P$ whose elements are $p_{ij}$ is called the
**transition probability matrix** of the process. $$P = \begin{bmatrix}
    p_{11} & p_{12} & p_{13} & \dots   \\
    p_{21} & p_{22} & p_{23} & \dots   \\
    p_{31} & p_{32} & p_{33} & \dots   \\
    \vdots & \vdots & \vdots  & \vdots \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix}.$$ Note that the elements of the matrix $P$ satisfies the
following properties:
$$\ 0 \le p_{ij} \le 1, \quad \text{ and } \quad \sum_{j \in S}p_{ij} = 1,$$
for all $i,j \in S.$ A matrix that satisfies these properties is called
a **stochastic matrix**.

::: {.example #NCD }
*No claims discount (NCD) policy: Let $X_n$ be the
discount status of a policyholder at time $n$. There are three levels of
discount, i.e. $S = \{0,1,2\}$ corresponding to three discount levels of
0, 20% and 40%. The following rules are assumed:*

-   *For a claim-free year, the policyholder moves up a level or remains
    in state 2 (the maximum discount state).*

-   *If there is at least one claim, the policyholder moves down one
    level or remains in state 0.*

*Suppose also that the probability of a claim-free year is $p$ and is
independent of $n$. The transition probability matrix is given by
$$P = \begin{bmatrix}
    1- p & p & 0    \\
    1-p & 0 & p   \\
    0 & 1-p & p    \\
    %\vdots & \vdots & \vdots  & \vdots \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix}.$$ The transition diagram is illustrated in the following
figure.*

*The following questions are of interest.*

1.  *What is the probability
    $$\Pr(X_0 = i_0, X_1 = i_1, X_2 = i_2, \ldots, X_n = i_n)?$$*

2.  *What is the probability
    $$\Pr(X_{n+1} = i_{n+1}, X_{n+2} = i_{n+2},  \ldots, X_{n+m} = i_{n+m}| X_n = i_n, X_{n-1} = i_{n-1}, \ldots, X_0 = i_{0})?$$*

3.  *What is the probability of transferring from state $i$ to state $j$
    in $n$ steps $$\Pr(X_{m+n} = j | X_m = i )?$$*

4.  *What is the long-term behavior of the Makov chain, i.e.
    $\lim_{n \rightarrow \infty} \Pr(X_n = j), j = 0,1,2$ given that
    $\Pr(X_0 = 0)$.*
:::

Later we will apply matrix algebra to compute these types of
probabilities and long-term probabilities.

::: {.example #healthInsurance }
*In a health insurance system, at the end of each day an
insurance company classifies policyholders as Healthy, Sick or Dead,
i.e. $S = \{H, S, D\}$. The following transition matrix $P$ for a
healthy-sick-dead model is given by $$P = \begin{bmatrix}
    p_{11} & p_{12} & p_{13}    \\
    p_{21} & p_{22} & p_{23}   \\
   0 & 0 & 1   \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix}.$$*
:::

The transition diagram is shown below.

It turns out that the probabilistic description of the Markov chain is
completely determined by its transition probability matrix and its
initial probability distribution $X_0$ at time 0.

::: {.example}
*By using the definition of conditional probabilities,
show that
$$\Pr(X_0 = i_0, X_1 = i_1, X_2 = i_2, \ldots, X_n = i_n) = \mu_{i_0}\, p_{i_0i_1} \cdots \, p_{i_{n-2} i_{n-1}}\, p_{i_{n-1} i_n},$$
where $\boldsymbol{\mu} = \boldsymbol{\mu}^{(0)}$ is the distribution of
initial random variable $X_0$, i.e. $\mu_i = \Pr(X_0 = i)$ (the
probability mass function of $X_0$).*
:::
**Solution:**
$$\begin{aligned}
 &\Pr(X_0 = i_0, X_1 = i_1, \ldots, X_n = i_n) \\
  &= \Pr(X_0 = i_0, X_1 = i_1, \ldots, X_{n-1} = i_{n-1}) \cdot \Pr(X_n = i_n | X_0 = i_0, X_1 = i_1, \ldots, X_{n-1} = i_{n-1})\\
  &= \Pr(X_0 = i_0, X_1 = i_1, \ldots, X_{n-1} = i_{n-1}) \cdot \Pr(X_n = i_n |  X_{n-1} = i_{n-1})\\
  &=  \Pr(X_0 = i_0, X_1 = i_1, \ldots, X_{n-1} = i_{n-1}) \cdot p_{i_{n-1} i_n} \\
  &\quad \vdots \\
  &= \mu_{i_0}\, p_{i_0i_1} \cdots \, p_{i_{n-2} i_{n-1}}\, p_{i_{n-1} i_n}.
  \end{aligned}$$

::: {.example #MKProperty1 }
*Show that $$\begin{aligned}
\Pr(X_{n+1} &= i_{n+1}, X_{n+2} = i_{n+2},  \ldots, X_{n+m} = i_{n+m}| X_n = i_n, X_{n-1} = i_{n-1}, \ldots, X_0 = i_{0}) \\
            &= \Pr(X_{n+1} = i_{n+1}, X_{n+2} = i_{n+2},  \ldots, X_{n+m} = i_{n+m}| X_n = i_n) \\
            &=  p_{i_{n} i_{n+1}} \cdots \, p_{i_{n+m-2} i_{n+m-1}}\, p_{i_{n+m-1} i_{n+m}}.\end{aligned}$$*
:::
**Solution:**
$$\begin{aligned}
&\Pr(X_{n+1} = i_{n+1}, X_{n+2} = i_{n+2},  \ldots, X_{n+m} = i_{n+m}| X_n = i_n, X_{n-1} = i_{n-1}, \ldots, X_0 = i_{0}) \\
&= \Pr(X_{n+1} = i_{n+1}, X_{n+2} = i_{n+2},  \ldots, X_{n+m} = i_{n+m}| X_n = i_n) \\ 
&= \Pr( X_{n+2} = i_{n+2},  \ldots, X_{n+m} = i_{n+m}| X_{n+1} = i_{n+1}, X_n = i_n)  \cdot \Pr(X_{n+1} = i_{n+1} |  X_n = i_n ) \\ 
&= \Pr( X_{n+2} = i_{n+2},  \ldots, X_{n+m} = i_{n+m}| X_{n+1} = i_{n+1})  \cdot p_{i_n i_{n+1}} \\ 
&\quad \vdots \\
&= \Pr( X_{n+m} = i_{n+m}| X_{n+m-1} = i_{n+m-1}) \cdots \Pr( X_{n+2} = i_{n+2}| X_{n + 1} = i_{n + 1}) \cdot p_{i_n i_{n+1}} \\ 
&= p_{i_{n+m-1} i_{n+m}} \cdot p_{i_{n+m-2} i_{n+m-1}}   \cdots p_{i_n i_{n+1}} .\end{aligned}$$

**Note** More general probabilities of the possible realisations of the process
can be calculated by summing the probabilities of elementary elements of
these forms.

::: {.example}
*For the NCD system defined on the state space
$S = \{0,1,2\}$ as given in Example \@ref(exm:NCD)
<!-- Example [Example 2](#NCD){reference-type="ref" reference="NCD"} -->
, suppose
that the probability of a claim-free year $p = 3/4$, and the
distribution of the initial discount rate
$\boldsymbol{\mu}  = (0.5,0.3,0.2)$. Find the following:*
:::

1.  $\Pr(X_0 = 2, X_1 = 1, X_2 = 0).$

2.  $\Pr( X_1 = 1, X_2 = 0 | X_0 = 2).$

3.  $\Pr(X_{10} = 2, X_{11} = 1, X_{12} = 0).$

4.  $\Pr( X_{11} = 1, X_{12} = 0 | X_{10} = 2).$

**Solution:**
The corresponding transition matrix is $$P = \begin{bmatrix}
    1/4 & 3/4 & 0    \\
    1/4 & 0 & 3/4   \\
    0 & 1/4 & 3/4    \\
    %\vdots & \vdots & \vdots  & \vdots \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix}.$$

1.  Denote $\boldsymbol{\mu}  = (\mu_1, \mu_2, \mu_3) =  (0.5,0.3,0.2)$
    $$\begin{aligned}
        \Pr(X_0 = 2, X_1 = 1, X_2 = 0) &=  \Pr(X_0 = 2, X_1 = 1) \cdot \Pr(X_2 = 0 | X_0 = 2, X_1 = 1) \\
        &= \Pr(X_0 = 2, X_1 = 1) \cdot \Pr(X_2 = 0 | X_1 = 1) \quad \text{(by Markov property)}\\
        &= \Pr(X_0 = 2) \cdot \Pr(X_1 = 1 | X_0 = 2 ) \cdot  \Pr(X_2 = 0 | X_1 = 1) \quad  \text{(again by Markov property)}\\
        &= \mu_3 p_{32} p_{21} = 0.2\cdot(1/4)\cdot(1/4) = 1/80.\end{aligned}$$

    Alternatively, it follows from
    Example \@ref(exm:MKProperty1)
    that,
    $$\Pr(X_0 = 2, X_1 = 1, X_2 = 0) = \mu_3 p_{32} p_{21} = 0.2\cdot(1/4)\cdot(1/4) = 1/80.$$

2.  $$\begin{aligned}
    \Pr( X_1 = 1, X_2 = 0 | X_0 = 2)
    &=   \Pr( X_1 = 1 | X_0 = 2)  \cdot \Pr( X_2 = 0 | X_1 = 1, X_0 = 2) \\ 
    &=   \Pr( X_1 = 1 | X_0 = 2) \cdot \Pr( X_2 = 0 | X_1 = 1) \quad \text{(by Markov property)}\\ 
    &= p_{32} p_{21} = (1/4)\cdot(1/4) = 1/16.\end{aligned}$$

3.  Following conditional probability, the Markov property, and
    time-homogeneity (to be discussed later) results in
    $$\begin{aligned}
        \Pr(X_{10} = 2, X_{11} = 1, X_{12} = 0) &=  \Pr(X_{10} = 2)  \cdot \Pr(X_{11} = 1, X_{12} = 0 | X_{10} = 2) \\
         &=  \Pr(X_{10} = 2)  \cdot \Pr(X_{11} = 1 | X_{10} = 2) \cdot \Pr( X_{12} = 0 | X_{10} = 2, X_{11} = 1)  \\
         &=  \Pr(X_{10} = 2)  \cdot \Pr(X_{11} = 1 | X_{10} = 2) \cdot \Pr( X_{12} = 0 |  X_{11} = 1)  \\
         &= \Pr(X_{10} = 2)  \cdot \Pr(X_{1} = 1 | X_{0} = 2) \cdot \Pr( X_{1} = 0 |  X_{0} = 1) \\
          &= \Pr(X_{10} = 2)  \cdot p_{32} p_{21} = 0.6922\cdot (1/4)\cdot(1/4) = 0.0433.  \\\end{aligned}$$
    Later we will show that
    $\Pr(X_{10} = 2) = (\boldsymbol{\mu} P^{10})_3 = 0.6922$ (here
    $(\boldsymbol{\mu} P^{10})_i$ denotes the $i$-th entry of the vector
    $\boldsymbol{\mu} P^{10}$.

4.  From conditional probability, the Markov property, and
    time-homogeneity, it follows that $$\begin{aligned}
    \Pr( X_{11} = 1, X_{12} = 0 | X_{10} = 2)
    &=   \Pr( X_{11} = 1 | X_{10} = 2)  \cdot \Pr( X_{12} = 0 | X_{11} = 1, X_{10} = 2) \\ 
    &=   \Pr( X_{11} = 1 | X_{10} = 2) \cdot \Pr( X_{12} = 0 | X_{11} = 1) \quad \text{(by Markov property)}\\ 
    &=   \Pr( X_{1} = 1 | X_{0} = 2) \cdot \Pr( X_{1} = 0 | X_{0} = 1) \quad \text{(by time-homogeneity)}\\ 
    &= p_{32} p_{21} = (1/4)\cdot(1/4) = 1/16.\end{aligned}$$

<!-- The Chapman-Kolmogorov equations -->
<!-- --------------------------------- -->

##  The Chapman-Kolmogorov equation and $n$-step transition probabilities


The $n$-step transition probability denoted by $p^{(n)}_{ij}$ is the
probability that the process goes from state $i$ to state $j$ in $n$
  transitions, i.e. $$p^{(n)}_{ij} = \Pr(X_{m+n} = j | X_m = i ).$$ Note
that for homogeneous process, the left hand side of the above equation
does not depend on $m$. Suppose that the transition from state $i$ at
time $m$ to state $j$ at time $m+n$ (i.e. in $n$ steps), going via state
$k$ in $l$ steps. One needs to examine all possible paths (from $i$ to
                                                           $k$ and then $k$ to $j$) and hence the $n$-step transition probability
$p^{(n)}_{ij}$ can be expressed as the sum of the product of the
transition probabilities $p^{(l)}_{ik} \, p^{(n-l)}_{kj}$.
$$\begin{aligned}
p^{(n)}_{ij} = \sum_{k \in S}p^{(l)}_{ik} p^{(n-l)}_{kj}, \quad 0 < l < n  \end{aligned}$$
  To derive the Chapman-Kolmogorov equation, we proceed as follows:
  
  $$\begin{aligned}
p^{(n)}_{ij} &= \Pr(X_n = j | X_0 = i) \\ 
&= \sum_{k \in S} \Pr(X_n = j , X_l = k | X_0 = i)    \\
&= \sum_{k \in S} \Pr(X_n = j | X_l = k , X_0 = i)  \cdot   \Pr(X_l = k |  X_0 = i) \\
&= \sum_{k \in S} \Pr(X_n = j | X_l = k )  \cdot   \Pr(X_l = k |  X_0 = i) \\
&= \sum_{k \in S}p^{(n-l)}_{kj} p^{(l)}_{ik} , = \sum_{k \in S}p^{(l)}_{ik} p^{(n-l)}_{kj}, \quad 0 < l < n . \end{aligned}$$
  
  This result is known as the Chapman-Kolmogorov equation. This relation
can be expressed in terms of matrix multiplication as
$$P^n  = P^l P^{n-l}.$$ The $n$-**step transition probabilities**
  $p^{(n)}_{ij}$ are the $ij$ elements of $P^n$.

::: {.example}
*For the NCD system given in Example \@ref(exm:NCD),
<!-- Example [Example 2](#NCD){reference-type="ref" reference="NCD"},  -->
  suppose that $p = 3/4$, the probability of a claim-free year and the initial
  discount level of a policyholder is 1 (with 20% discount).*

1. *Calculate the probability of starting with a discount level of 20% and ending up 3 years later at the same level.*
2. *Calculate the policyholder's expected level of discount after 3 years.*
:::
**Solution:**
The transition matrix is $$P = \begin{bmatrix}
    1/4 & 3/4 & 0    \\
    1/4 & 0 & 3/4   \\
    0 & 1/4 & 3/4    \\
    %\vdots & \vdots & \vdots  & \vdots \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix},$$ and $$P^3 = \begin{bmatrix}
    7/64 & 21/64 & 9/16    \\
    7/64 & 3/16 & 45/64   \\
    1/16 & 15/64 & 45/64    \\
    %\vdots & \vdots & \vdots  & \vdots \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix}.$$

1.  The probability of starting with a discount level of 20% and ending
    up 3 years later at the same level is equal to
    $p_{11}^{(3)} = 3/16,$ which is the element in the second row and
    second column of the matrix $P^3$ (not to be confused with the
    indices used) .

2.  The policy's expected level of discount after 3 years is
  $$\begin{aligned}
  \mathrm{E}[X_3 | X_0 = 1] &= \sum_{j=0}^{2} j \cdot  \Pr(X_3 = j | X_0 = 1) \\
  &= 0 \cdot (7/64) + 1 \cdot (3/16) + 2 \cdot (45/64)   \\
  &= 51/32 = 1.59375.\end{aligned}$$
    

##  Distribution of $X_n$

Let $\boldsymbol{\mu}^{(n)}$ be the vector of probability mass function
of $X_n$, i.e. $$\boldsymbol{\mu}^{(n)} = (\mu_1, \mu_2, \ldots ),$$
where $\mu_i = \Pr(X_n = i)$. It follows that
$$\boldsymbol{\mu}^{(n+1)} = \boldsymbol{\mu}^{(n)} P$$ and, in general,
$$\boldsymbol{\mu}^{(n+m)} = \boldsymbol{\mu}^{(n)} P^m.$$

::: {.example}
Consider the following questions:

1. Show that
$$Pr(X_1 = i) = \sum_{k \in S} \mu_k p_{ki} = ( \boldsymbol{\mu} P)_i,$$
  which is the $i$th element of the vector $\boldsymbol{\mu} P.$ Here
$\boldsymbol{\mu} = \boldsymbol{\mu}^{0}$ is the distribution of
initial random variable $X_0$ with $\mu_i = \Pr(X_0 = i)$.

2. In general, show that the distribution of $X_n$ is given by
$$Pr(X_n = i) =  ( \boldsymbol{\mu} P^n)_i.$$
:::
**Solution:**



1.  $$\begin{aligned}
        \Pr(X_1 = i) &= \sum_{k \in S} \Pr(X_1 = i | X_0 = k) \cdot \Pr(X_0 = k) \\
        &=  \sum_{k \in S} \mu_k \cdot p_{ki} \\
        &= ( \boldsymbol{\mu} P)_i.\end{aligned}$$

::: {.example #weather}
*The simple weather pattern can be classified into three
types including rainy ($R$), cloudy ($C$) and sunny ($S$). The weather
is observed daily. The following information is provided.*

-   *On any given rainy day, the probability that it will rain the next
    day is 0.7; the probability that it will be cloudy the next day
    0.2.*

-   *On any given cloudy day, the probability that it will rain the next
    day is 0.75; the probability that it will be sunny the next day
    0.1.*

-   *On any given sunny day, the probability that it will rain the next
    day is 0.2; the probability that it will be sunny the next day 0.4.*

*The weather forecast for tomorrow shows that there is a 40% chance of
rain and a 60% chance of cloudy. Find the probability that it will sunny
2 days later.*
:::
**Solution:**
As the ordered state of the chain is $R, C, S$, the initial distribution
is $\boldsymbol{\mu} = (0.4, 0.6, 0)$. The transition matrix $P$ is
given by $$P = \begin{bmatrix}
    0.7 & 0.2 & 0.1    \\
    0.75 & 0.15 & 0.1   \\
    0.2 & 0.4 & 0.4   \\
    %\vdots & \vdots & \vdots  & \vdots \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix},$$ and $$P^2 = \begin{bmatrix}
    0.66 & 0.21 & 0.13    \\
    0.6575 & 0.2125 & 0.13   \\
    0.52 & 0.26 & 0.22   \\
    %\vdots & \vdots & \vdots  & \vdots \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix}.$$ This gives
$$\boldsymbol{\mu} \cdot P^2 = (0.6585,0.2115, 0.13).$$ Hence, the
desired probability of sunny is
$$\Pr(X_2 = S) =( \boldsymbol{\mu} \cdot P^2 )_S  = (\boldsymbol{\mu} \cdot P^2 )_3 = 0.13.$$

##  Joint Distribution

Let $X_0, X_1, \ldots$ be a Markov chain with transition matrix $P$ and
initial distribution $\boldsymbol{\mu}.$ For all
$0 \le n_1 \le n_2 < \cdots < n_{k-1} < n_k$ and states
$i_1, i_2, \ldots , i_{k-1}, i_k,$
$$P(X_{n_1} = i_1, X_{n_2} = i_2,\ldots, X_{n_k} = i_{k1}1, X_{n_k} = i_k)
= (\boldsymbol{\mu} P^{n_1} )_{i_1} (P^{n_2 - n_1} )_{i_1i_2} \cdots (P^{n_k - n_{k -1}} )_{i_{k-1}i_k}.$$
From the above result, the joint probability is obtained from just the
initial distribution $\boldsymbol{\mu}$ and the transition matrix $P$.

::: {.example}
*In Example \@ref(exm:weather), on Sunday, the chances of rain, cloudy and sunny
have the same probabilities. Find the probability that it will be sunny
on the following Wednesday and Friday, and cloudy on Saturday.*
:::
**Solution:**
We are given that $\boldsymbol{\mu} = (1/3, 1/3, 1/3).$ From
$$P^3 = \begin{bmatrix}
    0.645500 & 0.215500 & 0.139   \\
    0.645625 & 0.215375 & 0.139  \\
    0.603000 & 0.231000 & 0.166   \\
    %\vdots & \vdots & \vdots  & \vdots \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix},$$ the required probability is $$\begin{aligned}
 \Pr(X_3 = S, X_5 = S, X_6 = C) &= (\boldsymbol{\mu}  \cdot P^3)_S \cdot  P^{2}_{SS} \cdot P_{SC} \\
&= (\boldsymbol{\mu}  \cdot P^3)_3 \cdot P^{2}_{33} \cdot P_{32}  \\
&= 0.148 \cdot  0.22 \cdot 0.4 = 0.013024.\end{aligned}$$

##  Random walk with absorbing and reflecting barrier(s)

::: {.example}
*A one-dimensional random walk $\{X_n\}$ is defined on a
finite or infinite subset of integers in which the process in state $i$
can either stay in $i$ or move to its neighbouring states $i -1$ and
$i+1$. Suppose that given that $X_n = i$ at time $n$,*

-   *the probability of moving to state $i+1$ is $p_i$,*

-   *the probability of remaining in state $i$ is $r_i$, and*

-   *the probability of moving to state $i-1$ is $q_i$,*

*where $p_i + q_i + r_i = 1$ for all $i$.*

1.  *Write down the transition matrix.*

2.  *Show that the random walk has Markov property.*
:::
**Solution:**
1.  The transition diagram and the transition matrix are infinite:

$$P = \begin{bmatrix}
        \ddots & \cdots & \cdots & \cdots & \cdots & \cdots  & \cdots    \\
        \cdots & q_{-1} & r_{-1}  & p_{-1}  & \cdots & \cdots  & \cdots  \\
        \cdots & \cdots & q_{0} & r_{0}  & p_{0}  & \cdots   & \cdots  \\
        \cdots & \cdots & \cdots & q_{1} & r_{1}  & p_{1}    & \cdots    \\
        \cdots & \cdots & \cdots & \cdots & \cdots  & \cdots  & \ddots   \\    
     %0 & 0 & 0 & 1/2 & 1/2     \\
        %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
    \end{bmatrix}.$$

2.  Clearly, the Markov property holds because $$\begin{aligned}
        &\Pr(X_{n+1} = k | X_n = i, X_{n-1} = i_{n-1}, \ldots, X_0 = i_0)  \\
        &= \Pr(X_{n+1} = k | X_n = i) =
        \begin{cases}
                   p_i,              & k = i+1\\
                   r_i,              & k = i\\
                   q_i,               & k = i-1\\               
                   0,       & \text{otherwise}.
               \end{cases}\end{aligned}$$

::: {.example #simpleRW}
*The random walk can be used to model the fortune of a
gambler. The gambler bets per game and the probability of winning is $p$
and the probability of losing is $q$ where $p + q = 1$. In addition, the
gambler is ruined (or goes broke) if he reaches state 0, and also stops
the game if he reaches state $N$. Therefore, the state space is
$S = \{0, 1, \ldots, N\}$. Note that
$$p_{00} = 1 \text { and } p_{NN} = 1.$$ The states $0$ and $N$ are
referred to as **absorbing boundaries (absorbing states)** and the
remaining states $1,2,\ldots,N-1$ are **transient**. Roughly speaking,
if a state is known as transient if there is a possibility of leaving
the state and never returning.*
:::
**Solution:**
The transition diagram and the transition matrix of the simple random
walk with absorbing boundaries (states) are given as follows:

$$P = \begin{bmatrix}
    1& 0 & 0 & 0& \cdots & 0& 0  & 0    \\
    q & 0 & p  & 0& \cdots & 0& 0  & 0  \\
    0 & q & 0 & p  & \cdots & 0& 0 & 0   \\
    \vdots & \vdots & \vdots & \vdots &  \ddots & \vdots    & \vdots & \vdots   \\
    0 & 0 & 0 & 0  & \cdots & q & 0 & p   \\
    0 & 0 & 0 & 0  & \cdots & 0 & 0 & 1   \\
 %0 & 0 & 0 & 1/2 & 1/2     \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix}.$$

1.  In general, a state $i$ is called **absorbing** if $p_{ii} = 1$

2.  The simple random walk as given in Example \@ref(exm:simpleRW) can be modified so that whenever the process is in state 0 (or state
    $N$),

    -   the probability of remaining in state 0 is $\alpha$, and

    -   the probability of moving to state 1 is $1 - \alpha$.

    In this case, the state 0 is referred to as a **reflecting barrier**
    for the chain. The process might be used to model the fortune of an
    individual when negative fortune is reset to zero.

##  An example of nonhomogeneous Markov chain


In this section, we give an example of a discrete-time nonhomogeneous
Markov chain. Again, without stated otherwise, we shall assume that the
discrete-time Markov chains are homogeneous.

::: {.example}
*(Adapted from W.J.Stewart)
A Markov chain $X_0, X_1, \ldots$ consists of two states $\{1,2\}$. At
time step $n$, the probability that the Markov chain remains in its
current state is given by $$p_{11}(n) = p_{22}(n) = 1/n,$$ while the
probability that it changes state is given by
$$p_{12}(n) = p_{21}(n) = 1 - 1/n.$$*

1.  *Draw a transition diagram of the Markov chain.*

2.  *Write down the transition matrix.*

3.  *Calculate $\Pr(X_5 = 2, X_4 = 2, X_3 = 1, X_2 =1 | X_1 = 1)$.*
:::
**Solution:**
1.  The transition diagram and the transition matrix are dependent of
    the time step $n$, and are given as follows:

  $$P(n) = \begin{bmatrix}
        \frac{1}{n} & \frac{n-1}{n}   \\
        \frac{(n-1)}{n} & \frac{1}{n}   \\
        %\vdots & \vdots & \vdots  & \vdots \\
        %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
    \end{bmatrix}.$$

2.  The probability of taking a particular part can be calculated by
    $$\begin{aligned}
        \Pr(X_5 = 2, X_4 = 2, X_3 = 1, X_2 =1 | X_1 = 1)  &= p_{11}(1) \cdot p_{11}(2) \cdot p_{12}(3) \cdot p_{22}(4) \\
            &= 1 \cdot 1/2 \cdot 2/3 \cdot 1/4 = 1/12.\end{aligned}$$
    Other paths lead to state 2 after four transitions, and have
    different probabilities according to the route they follow. What is
    important is that, no matter which route is chosen, once the Markov
    chain arrives in state 2 after four steps, the future evolution is
    specified by $P(5)$, and not any other $P(i), i \le 4$.



## Simulation


Simulation is a powerful tool for studying Markov chains. For many
Markov chains in real-world applications, state spaces are large and
matrix methods may not be practical.

A Markov chain can be simulated from an initial distribution and
transition matrix. To simulate a Markov sequence $X_0,X_1, \ldots,$
simulate each random variable sequentially conditional on the outcome of
the previous variable. That is, first simulate $X_0$ according to the
initial distribution. If $X_0 = i$, then simulate $X_1$ from the $i-th$
row of the transition matrix. If $X_1 = j$, then simulate $X_2$ from the
$j-th$ row of the transition matrix, and so on.

    <!-- [frame=single, escapeinside={(*}{*)}, caption={Algorithm for Simulating a Markov Chain}] -->
    
    Algorithm for Simulating a Markov Chain
    
    Input: (i) initial distribution  (*$\boldsymbol{\mu}$*), (ii) transition matrix (*$P$*), (iii) number of steps (*$n$*).
    Output: (*$X_0, X_1, \ldots , X_n$*)
    Algorithm:
        Generate (*$X_0$*) according to (*$\boldsymbol{\mu}$*)
        FOR (*$i = 1, \ldots ,n$*)
            Assume that (*$X_{i-1} = j$*)
            Set (*$\boldsymbol p = j-$*)th row of (*$P$*)
            Generate (*$X_i$*) according to (*$\boldsymbol p$*)
        END FOR

##  Monte Carlo Methods

Monte Carlo methods are simulation-based algorithms that rely on
generating a large set of samples from a statistical model to obtain the
behaviour of the model and estimate the quantities of interest. For a
large sample set of a random variable representing a quantity of
interest, the law of large numbers allows to approximate the expectation
by the average value from the samples.

Consider repeated independent trials of a random experiment. We will
need to generate a large number of samples $X_1, X_2, \ldots$ from the
model. A Monte Carlo method for estimating the expectation
$\mathrm{E}( X )$ is a numerical method based on the approximation
$$\mathrm{E}(X) \approx \frac{1}{N}\sum_{i=1}^N X_i,$$ where
$X_1, X_2, \ldots$ are i.i.d. with the same distribution as $X$.

While computing expectations and computing probabilities at first look
like different problems, the latter can be reduced to the former: if $X$
is a random variable, we have
$$\Pr(X \in A) = \mathrm{E}(1_A(X)).$$

Using this equality, we can estimate $\Pr(X \in A)$ by
$$\Pr(X \in A) = \mathrm{E}(1_A(X)) = \frac{1}{N}\sum_{i=1}^N 1_A(X_i).$$

Recall that the indicator function of the set $A$ is the defined as 

$$\begin{aligned}
        1_A(x) &= 
        \begin{cases}
                   1,  & \text{if } x \in A\\               
                   0,       & \text{otherwise}.
               \end{cases}\end{aligned}$$


The following user-defined function in Excel can be used to simulate
random numbers from a discrete distribution.

    <!-- [frame=single, escapeinside={(*}{*)}, caption={A user-defined function in Excel to simulate random numbers from a discrete distribution.  -->
    <!-- }] -->

<div class="verbatim">
    A user-defined function in Excel to simulate random numbers from a discrete distribution.
    Public Function Discrete(value As Variant, prob As Variant)

    Dim i As Integer
    Dim cumProb As Single
    Dim uniform As Single
    Randomize
    'Randomize Statement
    'Initializes the random-number generator.

    Application.Volatile

    ' This example marks the user-defined function Discrete as volatile.
    ' The function will be recalculated when any cell in any workbook
    ' in the application window changes value worksheet.

    uniform = Rnd
    cumProb = prob(1)
    i = 1
    Do Until cumProb > uniform
        i = i + 1
        cumProb = cumProb + prob(i)
    Loop
    Discrete = value(i)

    End Function
</div>





```{r, include=FALSE}
tutorial::go_interactive()
```

<script src=https://cdn.datacamp.com/datacamp-light-latest.min.js></script>

```{r MC,  tut = FALSE, echo = TRUE}
library(expm)
library(markovchain)
library(diagram)
library(pracma)

stateNames <- c("Rain","Nice","Snow")
Oz <- matrix(c(.5,.25,.25,.5,0,.5,.25,.25,.5),
             nrow=3, byrow=TRUE)
row.names(Oz) <- stateNames; 
colnames(Oz) <- stateNames
Oz



plotmat(Oz,pos = c(1,2),
        lwd = 1, box.lwd = 2,
        cex.txt = 0.8,
        box.size = 0.1,
        box.type = "circle",
        box.prop = 0.5,
        box.col = "light yellow",
        arr.length=.1,
        arr.width=.1,
        self.cex = .4,
        self.shifty = -.01,
        self.shiftx = .13,
        main = "")


Oz3 <- Oz %^% 3
round(Oz3,3)


u <- c(1/3, 1/3, 1/3)
round(u %*% Oz3,3)
```



We can use R to generate sample paths of a Markov chain. We first load the library `markovchain` package. See [https://cran.r-project.org/web/packages/markovchain/vignettes/an_introduction_to_markovchain_package.pdf](https://cran.r-project.org/web/packages/markovchain/vignettes/an_introduction_to_markovchain_package.pdf) for more details. 


<script src=https://cdn.datacamp.com/datacamp-light-latest.min.js></script>


```{r MC2}
library(markovchain)
weatherStates <- c("sunny", "cloudy", "rain") 
byRow <- TRUE
weatherMatrix <- matrix(data = c(0.70, 0.2, 0.1,
                                 0.3, 0.4, 0.3,
                                 0.2, 0.45, 0.35), 
                                 byrow = byRow, nrow = 3, dimnames = list(weatherStates, weatherStates))

mcWeather <- new("markovchain", states = weatherStates, byrow = byRow, transitionMatrix = weatherMatrix, name = "Weather")


# rmarkovchain: Function to generate a sequence of states from homogeneous or non-homogeneous Markov chains.
weathersOfDays <- rmarkovchain(n = 365, object = mcWeather, t0 = "sunny")
weathersOfDays[1:30]

weathersPaths <- as.data.frame(replicate(5,rmarkovchain(n = 365, object = mcWeather, t0 = "sunny") ))
```




The following user-defined function in Excel can be used to simulate
random numbers from a discrete distribution.

    [frame=single, escapeinside={(*}{*)}, caption={A user-defined function in Excel to simulate random numbers from a discrete distribution. 
    }]
    Public Function Discrete(value As Variant, prob As Variant)

    Dim i As Integer
    Dim cumProb As Single
    Dim uniform As Single
    Randomize
    'Randomize Statement
    'Initializes the random-number generator.

    Application.Volatile

    ' This example marks the user-defined function Discrete as volatile.
    ' The function will be recalculated when any cell in any workbook
    ' in the application window changes value worksheet.

    uniform = Rnd
    cumProb = prob(1)
    i = 1
    Do Until cumProb > uniform
        i = i + 1
        cumProb = cumProb + prob(i)
    Loop
    Discrete = value(i)

    End Function








::: {.example}
*(R or Excel) A gambler starts with and plays a game
where the chance of winning each round is 60%. The gambler either wins
or loses on each round. The game stops when the gambler either gains or
goes bust.*

1.  *Develop an Excel worksheet or create an R code to simulate 50 steps
    of the finite Markov chain of the random walk $X_n$ given in
    Example \@ref(exm:simpleRW). Repeat the simulation 10 times. How many of
    your simulations end at 0.*

2.  *Use the results from the simulations to estimate the mean and
    variance of $X_{5}$.*

3.  *Use the results from the simulations to estimate the probability
    that the gambler is eventually ruined.*
:::

::: {#exampleStationary2 .example}
*(R or Excel) A Markov chain $X_0, X_1, \ldots$ on
states $\{1,2\}$ has the following transition matrix
$$P = \begin{bmatrix}
    1-a & a   \\
    b & 1-b   \\
    %\vdots & \vdots & \vdots  & \vdots \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix},$$ 
where $0 < a,b < 1.$*

1.  *Use either Excel or R to estimate the long-term distribution of the
    Markov chain. (Hint: consider the $n$-step transition matrix for
    several increasing values of $n$). Comments on the results
    obtained.*

**Note** 
    *Later we will see that in many cases, a Markov chain exhibits a
    long-term limiting behaviour. The chain settles down to an
    equilibrium distribution, which is independent of its initial
    state.*

2.  *Use simulations to estimate the long-term probability that a Markov
    chain hits each of the states. (Hint: simulate the Markov chain 1000
    steps and calculate the proportion of visits to each state)*
:::

## Classification of states

Throughout this section $\{ X_n\}_{n \ge 0}$ is a time homogeneous
Markov chain with state space $S$ and transition matrix
$P = (p_{ij})_{i,j \in S}$.

For any $i, j \in S$,

-   The state $j$ can be **reached** from the state $i$, denoted by
    $i \rightarrow j$ if there is a nonzero probability
    $p^{(n)}_{ij} > 0$ for some $n \ge 0$.

-   The states $i$ and $j$ are said to **communicate**, or to be **in
    the same class**, and denoted by $i \leftrightarrow j$, if
    $i \rightarrow j$ and $j \rightarrow i$.

**Note** 
Note that $i \rightarrow j$ if and only if there exist states
$k_1, k_2, \ldots, k_r$ such that
$$p_{ik_1} p_{k_1k_2} \ldots p_{k_r j}>0,$$ i.e. it is not necessary
that $j$ can be reached from the state $i$ in one single step.

-   The relation $\leftrightarrow$ is an equivalence relation and
    partition the state space $S$ into equivalence classes, which are
    known as **classes (or communication classes)** of the Markov chain.
    Thus in any class all the states communicate, but none of them
    communicates with any state outside the class.

Additional properties for a communication class are defined as follows:

-   The class $C$ is said to be **closed** if it is impossible to reach
    any state outside $C$ from any state in $C$, i.e. escape from $C$ is
    impossible. Otherwise, the class $C$ is said to be **non-closed**,
    i.e. escape from $C$ is possible.

```{=html}
<!-- -->
```
-   If the entire state space $S$ is only one communication class (all
    states communicate), then it is necessarily closed and the Markov
    chain is said to be **irreducible**. Otherwise, the Markov chain is
    said to be **reducible**.

-   A closed class consisting of a single state is an **absorbing
    state**.

::: {.example}
*Consider each of the following Markov chains:*

(a) *NCD system (Example \@ref(exm:NCD)),*

(b) *the health insurance system (Example \@ref(exm:healthInsurance)), and*

(c) *a simple random walk (Example \@ref(exm:simpleRW)),*

*Identify the communication classes. Is the Markov chain irreducible?*
:::
**Solution:**
It is a good practice to use transition diagram and also verify the
answers.

(a) Every two states are intercommunicating, so $\{0,1,2\}$ is a single
    closed class, and hence the Markov chain is irreducible. This is
    because $0 \rightarrow 1$ and $1 \rightarrow 0$ ($p_{01} > 0$ and
    $p_{10} > 0$), and $1 \rightarrow 2$ and $2 \rightarrow 1$
    ($p_{12} > 0$ and $p_{21} > 0$).

(b) There are two classes of intercommunicating states, $O = \{H,S\}$ is
    non-closed, and $C = \{ D \}$ is closed (and also an absorbing
    state). Clearly, the Markov chain is not irreducible. This is
    because

    -   $p_{DD} = 1$, i.e. $C$ is a class.

    -   $O$ is open class because $p_{HS} > 0$, $p_{SH} >0$, so this is
        a class, and for example $p_{HD}$ \> 0 but $p^{(n)}_{DH} = 0$
        for all $n$ (i.e. one cannot leave $C$ starting from the state
        $D$), so O is an open class.

(c) The simple random walk with absorbing boundaries has three classes,
    $\{1,2, \ldots, N-1 \}$ is non-closed class, $\{0 \}$ and $\{N \}$
    are two closed classes.

::: {.example #exampleMC}
*A Markov chain with state space $S = \{1,2,3,4,5\}$ has
the following transition matrix: $$P = \begin{bmatrix}
    1 & 0 & 0 & 0 & 0     \\
    1/5 & 1/5 & 1/5 & 1/5 & 1/5   \\
    1/3 & 1/3 & 0 & 1/3 & 0     \\  
0 & 0 & 0 & 0 & 1     \\
0 & 0 & 0 & 1/2 & 1/2     \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix}.$$*

1.  *Draw a transition diagram.*

2.  *Identify the communication classes. Is the Markov chain
    irreducible?*
:::
**Solution:**
There are two closed classes $C_1 = \{1\}$ and $C_2 = \{4,5\}$ and one
non-closed class $O = \{ 2,3\}$. This is because

-   $C_1$ is closed because $p_{11} = 1$.

-   $C_2$ is a class because $4 \rightarrow 5$ ($p_{45} >0$) and
    $5 \rightarrow 4$ ($p_{54} >0$), and is closed because $p_{ij} = 0$
    for all $i \in C_2$ and $j \not\in C_2$.

-   $O$ is a class because $2 \rightarrow 3$ ($p_{23} >0$) and
    $3 \rightarrow 2$ ($p_{32} > 0$), and is non-closed because
    $p_{21} > 0$, but $p_{11} = 1$.

## Absorption probabilities and expected time to absorption


For the random walk with absorbing boundaries (i.e. $0$ and $N$), two
questions arises, in which state, $0$ or $N$ is the process eventually
absorbed (or trapped) and on the average how long does it take to reach
one of these absorbing states? We first define the following terms which
applies to the random walk process with absorbing boundaries.

The time of absorption $T$ is defined as
$$T = \min\{ n \ge 0 | X_n = 0 \text{ or } X_n = N \}$$ and the
**probability of eventually absorption** in state 0 is given by
$$u_i = \Pr\{ X_T = 0 | X_0 = i  \}, \text{ for } i = 1,2,\ldots,N-1.$$
The **mean time to absorption** of the process is given by
$$\mathrm{E}[T |  X_0 = i ] \, \text{ for } i = 1,2,\ldots,N-1.$$

##  First step analysis


First step analysis allows us to evaluate quantities of interest from
the Markov chain, for e.g. the absorption probabilities and the mean
duration until absorption. The method is based on considering all
possibilities at the end of the first transition and then apply the law
of total probability to formulate equations involved all unknown
quantities. We illustrate how to use the first step analysis in the
following Markov chain.

::: {.example #eg_absorption}
*Consider the Markov chain with state space
$S = \{0,1,2\}$ and transition probability matrix given by
$$P = \begin{bmatrix}
    1 & 0 & 0    \\
    p_{10} & p_{11} & p_{12}   \\
   0 & 0 & 1   \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix}.$$*
:::
**Solution:**
The classes and types are as follows:

-   Two closed classes are $C_1 = \{0\}$ and $C_2 = \{2\}$.

-   $\{1\}$ is a non-closed class.

Let us consider the problem of evaluating the absorption probabilities.
For any closed class $C$, define
$$u^C_{i} = \Pr(\text{Markov chain eventually absorbed in } C | X_0 = i).$$
Clearly, the absorption probabilities also depend on the initial states.
A vector of absorption probabilities is then given by
$\mathbf{u}^C = (u^C_i)_{i \in S}$ We suppress the superscript $C$ and
simply write $u^C_{i} = u_i$ and $\mathbf{u}^C = \mathbf{u}$.

Consider the closed class $C_1 = \{0\}$. We have $$\begin{aligned}
    u_{0} &= \Pr(\text{Markov chain eventually absorbed in } C_1 | X_0 = 0) = 1, \\ 
    u_{2} &= \Pr(\text{Markov chain eventually absorbed in } C_1 | X_0 = 2) = 0, \\ 
    u_{1} &= \Pr(\text{Markov chain eventually absorbed in } C_1 | X_0 = 1) = u_1. \\   \end{aligned}$$
By considering the first transition from state $1$ to either state 0, 1
and 2, and using the Markov property, the law of total probability gives
$$\begin{aligned}
    u_1 &= \Pr(\text{Markov chain eventually absorbed in } C_1 | X_0 = 1) \\
    &= \sum_{k = 0}^2 \Pr(\text{Markov chain eventually absorbed in } C_1 | X_0 = 1, X_1 = k)   \Pr(X_1 = k | X_0 = 1) \\
    &= \sum_{k = 0}^2 \Pr(\text{Markov chain eventually absorbed in } C_1 | X_1 = k)    \Pr(X_1 = k | X_0 = 1) \\
    &=  (p_{10}) \cdot  u_0 + (p_{11}) \cdot  u_1 + (p_{12}) \cdot  u_2  \\ 
    &= (p_{10})\cdot 1 + (p_{11}) \cdot  u_1 +  (p_{12}) \cdot 0. \end{aligned}$$
Solving for $u_1$ gives
$$u_1 =  u^{C_1}_1 = \frac{p_{10}}{1 - p_{11}} = \frac{p_{10}}{p_{10} +  p_{12}}.$$

**Note** 
Similarly, we have $$\begin{aligned}
    u_0 &= p_{00} \cdot u_0 + p_{01} \cdot  u_1 +  p_{02} \cdot  u_2 \\
    u_1 &= p_{10} \cdot  u_0 + p_{11} \cdot  u_1 + p_{12} \cdot  u_2  \\ 
    u_2 &= p_{20} \cdot  u_0 + p_{21} \cdot  u_1 + p_{22} \cdot  u_2,  \\\end{aligned}$$
where the first and the last equations reduce to $u_0 = u_0$ and
$u_2 = u_2$, respectively. In general, for a closed class $C$, the
vector of absorption probabilities $\mathbf{u}$ satisfies the following
system of linear equations:

1.  $\mathbf{u} = P\mathbf{u}$ (here $\mathbf{u}$ is treated as a column
    vector),

2.  $u_{i} = \Pr(\text{Markov chain eventually absorbed in } C) | X_0 = i) = 1$
    for all $i \in C$, and

3.  $u_{i} = \Pr(\text{Markov chain eventually absorbed in } C) | X_0 = i) = 0$
    for all $i$ in any other close classes.

::: {#absorption .example}
*In this example, consider the closed class
$C_2 = \{2\}$. Find the absorption probabilities $u^{C_2}_0, u^{C_2}_1$
and $u^{C_2}_2$. Comment on these results.*
:::
**Solution:**
For the closed class $C_2$, we proceed in the same way as in the closed
class $C_1$. Let $\mathbf{u}=\mathbf{u}^{C_{2}} = (u_0,u_1,u_2)^T$ be
the vector of absorption probabilities in the closed class $C_{2}=\{2\}$
with $u_0 = 0$ and $u_2 =1$. It follows that $$\begin{aligned}
    u_1 &=  p_{10} \cdot  u_0 + p_{11} \cdot  u_1 + p_{12} \cdot  u_2  \\
    &=  p_{11} \cdot  u_1 + p_{12}.\end{aligned}$$ Hence,
$u_1 = \frac{p_{12}}{1-  p_{11}} =\frac{p_{12}}{p_{10}+ p_{12}}$. It
should be emphasised that
$$\mathbf{u}^{C_1} + \mathbf{u}^{C_2} = \mathbf{1} .$$

**Notes** 
1.  For any initial state $i$, the sum of the absorption probabilities
    over all closed classes is 1 (as verified in Example
    [Example 19](#absorption){reference-type="ref"
    reference="absorption"}). In particular, when a Markov chain has two
    closed classes $C_1$ and $C_2$,
    $\mathbf{u}^{C_2} = \mathbf{1} - \mathbf{u}^{C_1}$.

2.  In the case when $S$ is finite or when the set of states in
    non-closed classes is finite, the vector $\mathbf{u}$ is the unique
    solution of the above system of linear equations.

::: {.example}
*Consider the Markov chain defined in Example \@ref(exm:exampleMC).
Find the absorption probabilities in the closed class $C_1 = \{1\}$ and
$C_2 = \{4,5\}$.*
:::
**Solution:**
Here $C_{1}=\{1\}$ and $C_{2}=\{4,5\}$ are closed classes and
$0 = \{2,3\}$ is an open class.

Let $\mathbf{u}=\mathbf{u}^{C_{1}}$ be the vector of absorption
probabilities in the closed class $C_{1}=\{1\}$. Write
$\mathbf{u}= (u_1,u_2, \ldots,u_5)^T$ and $u_1 = 1$ and $u_4 =  u_5 =0$,
From $\mathbf{u}=P \cdot \mathbf{u}$,

$$\left(\begin{array}{c}u_{1} \\ u_{2} \\ \vdots \\ u_{5}\end{array}\right)=P\left(\begin{array}{c}u_{1} \\ u_{2} \\ \vdots \\ u_{5}\end{array}\right) \text {gives}$$
$$\begin{aligned}
    u_2 &= \frac{1}{5} + \frac{1}{5} u_2 + \frac{1}{5} u_3 \\
    u_3 &= \frac{1}{3} + \frac{1}{3} u_2.   \end{aligned}$$ Solving the
linear system for $u_2$ and $u_3$ yields $u_2 = 4/11$ and $u_3 = 5/11$.
Hence, the absorption probabilities in the closed class $C_1$ is
$$\mathbf{u}= (1,4/11,5/11,0,0)^T.$$ In addition, since there are two
closed classes,
$\mathbf{u}^{C_2} = \mathbf{1} - \mathbf{u}^{C_1} = (0,7/11,6/11,1,1)^T.$

##  The expected time to absorption


The expected time to absorption can be determined by analysing all
possibilities occurring in the first step. We again consider the process
defined in Example \@ref(exm:eg_absorption) on the set $\{0, 1, 2\}$ with the transition
matrix $$P = \begin{bmatrix}
    1 & 0 & 0    \\
    p_{10} & p_{11} & p_{12}   \\
   0 & 0 & 1   \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix}.$$ The time of absorption $T$ is defined as
$$T = \min\{ n \ge 0 | X_n = 0 \text{ or } X_n = 2 \}$$ and the mean
time to absorption of the process is given by
$v = \mathrm{E}[T |  X_0 = 1 ] .$

The following observations can be made:

1.  The absorption time $T$ is always at least 1.

2.  If either $X_1 = 0$ or $X_1 = 2$, then no further steps are
    required.

3.  If $X_1 = 1$, then the process is back at its starting point and on
    the average $v$ additional steps are required for absorption.

Weighting all these possibilities by their respective probabilities, we
obtain the following equation $$\begin{aligned}
v &= 1 + p_{10} \cdot 0 + p_{11} \cdot v  + p_{12} \cdot 0    \\
  &= 1 +  p_{11} \cdot v,\end{aligned}$$ which results in
$$v = \frac{1}{1 - p_{11}}.$$

::: {.example #eg_absorption2}
*Consider the Markov chain with state space
$S = \{0,1,2,3\}$ and transition probability matrix given by
$$P = \begin{bmatrix}
    1 & 0 & 0  & 0  \\
    p_{10} & p_{11} & p_{12} & p_{13}   \\
    p_{20} & p_{21} & p_{22} & p_{23}   \\    
   0 & 0 & 0 & 1   \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix}.$$ Let $T$ be the time of absorption defined by
$$T = \min\{ n \ge 0 | X_n = 0 \text{ or } X_n = 3 \}$$ and the
**absorption probabilities** given by
$$u_i = \Pr\{ X_T = 0 | X_0 = i  \}, \text{ for } i = 1,2$$ and the mean
time to absorption of the process is given by
$$v_i  = \mathrm{E}[T |  X_0 = 1 ] , \text{ for } i = 1,2.$$ Calculate
the absorption probabilities and the mean time to absorption.*
:::
**Solution:**
There are 2 closed classes including $C_1= \{ 0 \}$, and $C_2= \{ 3 \}$,
and one non-closed class $O = \{1,2 \}$. Here,
$$u_i = u_i^{C_1} = \Pr\{ X_T = 0 | X_0 = i  \} = \Pr(\text{Markov chain eventually absorbed in } C_1 | X_0 = i), \text{ for } i = 1,2$$
By conditioning on the first step from state $i$ and using the Markov
property, we have $$u_i^{C_1} = \sum_{j \in S} p_{ij} u_j^{C_1}.$$
Clearly, $u_0^{C_1} = 1$ and $u_3^{C_1} = 0$. In particular, we have
$$\begin{aligned}
    u_1 &= p_{10} \cdot 1 + p_{11} \cdot u_1 + p_{12} \cdot u_2 \\
    u_2 &= p_{20} \cdot 1 + p_{21} \cdot u_1 + p_{22} \cdot u_2, \\\end{aligned}$$
which can also be obtained from the matrix equation
$\mathbf{u} = P\mathbf{u}$, where $\mathbf{u} = (1, u_1, u_2, 0)^T$. The
solution to the system of linear equations is $$\begin{aligned}
    u_1 &= \frac{ p_{10} (p_{22} - 1) - p_{12} p_{20}  }{ p_{11}(-p_{22}) +  p_{11}  + p_{12} p_{21}  + p_{22} - 1}, \\
    u_2 &= \frac{(  p_{11} - 1) p_{20} - p_{10} p_{21}  }{    p_{11}(-p_{22})  + p_{11}  + p_{12} p_{21} + p_{22} -1  }  . \\\end{aligned}$$

Similarly, the mean time to absorption also depends on the starting
state. By the first step analysis, we have for
$v_i  = \mathrm{E}[T |  X_0 = 1 ]$, $$\begin{aligned}
    v_1 &=  1 + p_{11} \cdot v_1 + p_{12} \cdot v_2 \\
    v_2 &=  1 + p_{21} \cdot v_1 + p_{22} \cdot v_2. \\\end{aligned}$$
Here the absorption time $T$ is always at least 1. If either $X_1 = 0$
or $X_1 = 3$, then no further steps are required. On the other hand, if
$X_1 = 1$ or $X_1 = 2$, then the process will require additional steps,
and on the average, these are $v_1$ and $v_2$. Weighting these two
possibilities, i.e. whether $X_1 = 1$ or $X_1 = 2$, by their respective
probabilities and summing according to the law of total probability
result in the above system of equations.

Solving the equations for $v_1$ and $v_2$ give the mean time to
absorption $$\begin{aligned}
    v_1 &=  \frac{-p_{12}  + p_{22} -1}{p_{11}(-p_{22}) + p_{11} + p_{12}p_{21} + p_{22} - 1}, \\
    v_2 &=  \frac{p_{11}  - p_{21} -1}{p_{11}(-p_{22}) + p_{11} + p_{12}p_{21} + p_{22} - 1}. \\\end{aligned}$$

## The long-term distribution of a Markov chain

In this section, we present another important property concerning
limiting behaviour of $P^n$ as $n \rightarrow \infty$ and hence the
long-term distribution of a Markov chain satisfying some certain
conditions. In particular, some Markov chains will converge to an
equilibrium (limiting) distribution, which is independent of its initial
state.

We also assume that the Markov chain with **a single closed class** $S$.

##  Stationary and limiting distributions for a single closed class


### Stationary distributions {#stationary-distributions .unnumbered}

Throughout this section, we consider a Markov chain whose transition
probability matrix is $P$ and state space $S$ is a single close class.
Then $S$ is necessarily closed and hence irreducible.

A probability distribution $\boldsymbol{\pi} = (\pi)_{i \in S}$ on $S$
is **stationary** if the following conditions hold (here
$\boldsymbol{\pi}$ is a row vector):

1.  $\pi_j = \sum_{i \in S} \pi_i p_{ij}$ or equivalently
    $\boldsymbol{\pi}  P  =\boldsymbol{\pi}$,

2.  $\pi_j \ge 0$,

3.  $\sum_{j \in S} \pi_j = 1$.

```{=html}
<!-- -->
```

**Notes** 
1.  For any stationary distribution $\boldsymbol{\pi}$, for all
    $n \ge 1$, $$\boldsymbol{\pi}  P^n  =\boldsymbol{\pi}.$$ Therefore,
    if we take $\boldsymbol{\pi}$ as the initial probability
    distribution, i.e. $\Pr(X_0 = i) = \pi_i$, then then the
    distribution of $X_n$ is also $\boldsymbol{\pi}$ , i.e.
    $\Pr(X_n = i) = \pi_i$

2.  The probability distribution $\boldsymbol{\pi}$ is said to be an
    invariant probability distribution.

3.  The most important property concerning the stationary
    distribution(which will be made formal) is that it gives the
    **long-term (limiting) distribution** of a Markov chain. In
    addition, $\pi_j$ also gives **the long run mean fraction of time**
    that the process $\{X_n\}$ is in state $j$.

::: {.example #exampleStationary }
*A Markov chain $X_0, X_1, \ldots$ on states $\{1,2\}$
has the following transition matrix $$P = \begin{bmatrix}
    1-a & a   \\
    b & 1-b   \\
    %\vdots & \vdots & \vdots  & \vdots \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix},$$ where $0 < a,b < 1.$*

1.  *Show that $$P^n = \frac{1}{a+b} \begin{bmatrix}
        b & a   \\
        b & a   \\
        %\vdots & \vdots & \vdots  & \vdots \\
        %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
    \end{bmatrix} +
    \frac{(1-a-b)^n}{a+b} \begin{bmatrix}
        a & -a   \\
        -b & b   \\
        %\vdots & \vdots & \vdots  & \vdots \\
        %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
    \end{bmatrix}.$$*

2.  *Show that the stationary probability distribution is
    $$\boldsymbol{\pi} = \left( \frac{b}{a+b},  \frac{a}{a+b} \right)  .$$*

3.  *Show that
    $$\lim_{n \rightarrow \infty} p_{ij}^{(n)} = \pi_j > 0 , \text{ for }j \in \{1,2\}.$$*
:::
**Solution:** The solutions can be found from the tutorial.

### Proportion of Time in Each State {#proportion-of-time-in-each-state .unnumbered}

The limiting distribution provides the long-term behaviour of the Markov
chain, i.e. it is the long-term probability that a Markov chain hits
each state. In this section, it can be shown that it also gives the
long-term proportion of time that the chain visits each state. Let us
consider a Markov chain $X_0, X_1, \ldots$ whose transition probability
matrix is $P$ and its limiting distribution is $\boldsymbol{\pi}$. Note
that the limiting distribution for the Markov chain satisfies
$$\lim_{n \rightarrow \infty}  p^{(n)}_{ij} =  \pi_j.$$

For each state $j$, define indicator random variable $$I_k =
    \begin{cases*}
      1, & if $X_k = j$ \\
      0,  & otherwise,
    \end{cases*}$$ for $k = 0,1,\ldots$. Hence, the number of times that
the Markov chain visits $j$ in the first $n$ steps is given by
$\sum_{k =0}^{n-1} I_k$ and the expected long-term proportion of time
that the chain visits state $j$ given that its initial state is $i$ is
$$\begin{aligned}
    \lim_{n \rightarrow \infty} \mathrm{E}\left(\frac{1}{n}  \sum_{k =0}^{n-1} I_k \, |\, X_0 = i \right) &= 
    \lim_{n \rightarrow \infty}  \frac{1}{n}  \sum_{k =0}^{n-1} \mathrm{E}( I_k \, |\, X_0 = i ) \\
    &= \lim_{n \rightarrow \infty}  \frac{1}{n}  \sum_{k =0}^{n-1} \Pr(X_k = j | X_0 = i)\\
    &= \lim_{n \rightarrow \infty}  \frac{1}{n}  \sum_{k =0}^{n-1} p^{(k)}_{ij}\\
    &= \lim_{n \rightarrow \infty}  p^{(n)}_{ij} =  \pi_j.\end{aligned}$$

Here we use the fact that if the sequence of numbers converges to a
limit, i.e. $x_n \rightarrow x$ as $n \rightarrow \infty$, then the
sequence of partial averages also converges to that limit, i.e.
$(x_1 + x_2 + \cdots x_n)/n \rightarrow x$ as $n \rightarrow \infty$.
This result is known as Cesaro's lemma.

::: {#weatherExample2 .example}
*Recall from
Example \@ref(exm:weather),
the simple weather pattern can be classified into three types including
rainy ($R$), cloudy ($C$) and sunny ($S$). The weather is observed daily
and can be modelled by the Markov transition matrix
$$P = \begin{bmatrix}
    0.7 & 0.2 & 0.1    \\
    0.75 & 0.15 & 0.1   \\
   0.2 & 0.4 & 0.4   \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix}.$$ It can be shown that the stationary distribution and
also the limiting distribution of the Markov chain is
$$\boldsymbol{\pi} = (94/147, 32/147,    1/7)$$ which gives the
proportions of visits to rainy, cloudy and sunny states are 94/147,
32/147, 1/7, respectively.*
:::

### The method of finding the stationary distribution {#the-method-of-finding-the-stationary-distribution .unnumbered}

To find the stationary distribution, we simply solve the linear
equations $\boldsymbol{\pi}  P  =\boldsymbol{\pi}$ (note that one of the
equations can be discarded), together with the condition
$\sum_{j \in S} \pi_j = 1$.

::: {.example}
*For the NCD process in Example \@ref(exm:NCD), the Markov
chain has the following transition probability matrix
$$P = \begin{bmatrix}
    1- p & p & 0    \\
    1-p & 0 & p   \\
    0 & 1-p & p    \\
\end{bmatrix}.$$ Find the stationary probability distribution of this
chain.*
:::
**Solution:**
Denote the stationary probability distribution by
$\boldsymbol{\pi} =  (\pi_1,  \pi_2, \pi_3)$. From
$\boldsymbol{\pi}  P  =\boldsymbol{\pi}$ and
$\pi_1 +   \pi_2 +  \pi_3 = 1$,
$$(\pi_1,  \pi_2, \pi_3)  \begin{bmatrix}
    1- p & p & 0    \\
    1-p & 0 & p   \\
    0 & 1-p & p    \\
\end{bmatrix} =  (\pi_1,  \pi_2, \pi_3),$$ which is equivalent to
$$\begin{aligned}
(1- p)\pi_1 + (1- p)\pi_2 &= \pi_1 \\
p\pi_1 + (1- p)\pi_3 &= \pi_2 \\
p\pi_2 + p\pi_3 &= \pi_3 \end{aligned}$$

By discarding one of the equations and adding the condition that
$\pi_1 +   \pi_2 +  \pi_3 = 1$, one can solve for
$\pi_1,  \pi_2, \pi_3$:
$$\pi_1 = \frac{(p-1)^2}{p^2 - p + 1},  \quad \pi_2 = \frac{-p^2 + p}{p^2 - p + 1},  \quad \pi_3 = \frac{p^2}{p^2 - p + 1}.$$

**Note** 
1.  In the above two examples, it can be shown that
    $$\lim_{n \rightarrow \infty} p_{ij}^{(n)} = \pi_j > 0 , \text{ for }j \in S,$$
    or, in terms of the Markov chain $\{X_n\}$,
    $$\lim_{n \rightarrow \infty} \Pr(X_n = j | X_0 = i) = \pi_j > 0, \text{ for }j \in S.$$
    This means that in the long run (as $n\rightarrow \infty$), the
    probability of finding Markov chain in state $j$ is approximately
    $\pi_j$ **no matter in which state the chain began at time 0**. This
    property holds for some Markov chains which satisfy **\"certain
    conditions\"**.

##  Sufficient conditions for the long-run behaviour of a Markov chain


In what follows, we will establish a set of sufficient conditions for
the long-run behaviour of a Markov chain. Two important results are
stated without proof.

::: {.theorem}
**Theorem 1**. *A Markov chain with a finite state space has at least
one stationary probability distribution.*
:::

::: {.theorem}
**Theorem 2**. *An irreducible Markov chain with a finite state space
has a unique stationary probability distribution.*
:::

::: {.example #rwInfiniteS }
*The simple random walks $X_n$ on
$S = \{\ldots, -2, -1, 0,1,2, \ldots \}$ is defined as
$$X_n = X_0 + \xi_1 + \xi_2 + \ldots + \xi_n,$$ where the random
variables $\xi_j$ are independent identically distributed with
$$\Pr(\xi = 1) = p, \quad \Pr(\xi = -1) = 1- p.$$ We can check that this
Markov chain is irreducible. However, the state space $S$ is infinite.
It can be checked directly from the equations
$\boldsymbol{\pi}  P  =\boldsymbol{\pi}$ that there is **no** stationary
distribution (given as an exercise).*
:::
**Solution:**
First we know that the entries of a stationary distribution sum to one.
Suppose the contrary that there is a stationary distribution
$\boldsymbol{\pi}$ for the simple random walk. Then by spatial
invariance of the simple random walk, $\pi_i$ is constant for all
$i \in S$ and also $\sum_{i\in S}\pi_i = 1$, which is impossible because
$S$ is infinite. Hence, there is **no** stationary distribution for the
simple random walk.

##  Limiting distributions {#limiting-distributions}


One of the important properties of stationary distributions is that the
distribution of the Markov chain satisfying certain conditions converges
to the stationary distribution. This result provides the long-term
behaviour of the Markov chain. In order to state the main result of this
section, we need to introduce another concept, namely the period of a
state.

A state $i$ is said to be **periodic** with period $d > 1$ if a return
to $i$ is possible only in a number of steps that is a multiple of $d$.
Equivalently, the period $d$ is the greatest common divisor of all
integers $n$ for which $p^{(n)}_{ii} > 0$. If the greatest common
divisor is 1, the state has period 1 and is said to be **aperiodic**.

A Markov chain in which each state has period 1 is called **aperiodic**.
Most Markov chains in applications are aperiodic.

::: {.example}
*Is the NCD system in
Example \@ref(exm:NCD)
aperiodic?*
:::
**Solution:**
The entire state space of the NCD system is a single class, and is
necessarily closed. Note also that $p_{00} > 0$ (and also $p_{22} > 0$),
i.e. the state 0 (and state 2) has an arrow back to itself.
Consequently, it is aperiodic because a return to this state is possible
in any number of steps (or the system can remain in this state in any
length of time).

Similarly, a return to state 1 is possible in $2,3, \ldots$ steps.
Therefore, the NCD system is aperiodic.

**Notes** 
1.  Periodicity is a class property, i.e. all states in one class have
    the same period (or if $i \leftrightarrow j$, then $i$ and $j$ have
    the same period).

2.  The Markov chain $\{X_n\}_{n\ge 0}$ is aperiodic if and only if
    there exists some $n >0$ such that
    $$p^{(n)}_{ij} > 0 \text{ for all } i,j \in S.$$ Such Markov chain
    is also called **regular**.

::: {.example}
*In the random walk model on a finite state space
$S = \{0,1,\ldots, N\}$ with absorbing boundaries in
Example \@ref(exm:simpleRW), determine the period of each state.*
:::
**Solution:**
The transition diagram and the transition matrix of the simple random
walk with absorbing boundaries (states) are given as follows:

The simple random walk with absorbing boundaries has three classes,
$O = \{1, 2, \ldots, N - 1\}$ is non-closed class, $C_1 = {0}$ and
$C_2 = {N}$ are two closed classes. For each state $i$ in $O$,
$p^{(2n)}_{ii} > 0$ and $p^{(2n+1)}_{ii} = 0$ for $n = 1,2,\ldots$.
Therefore, the state $i$ in this open communication class has period 2.

On the other hand, $p_{00} = 1$ (and also $p_{NN} =  1$) and hence, the
states $0$ and $N$ are aperiodic because a return to each of these
states is possible in any number of steps (or the system can remain in
this state in any length of time).

::: {.example}
*In the random walk model on an infinite state space
$S = \{\ldots,-2,-1,0,1,2, \ldots \}$ in
Example \@ref(exm:rwInfiniteS), determine the period of each state.*
:::
**Solution:**
The entire state space is a single class. Note also that
$p^{(2n)}_{ii} > 0$ and $p^{(2n+1)}_{ii} = 0$ for $n = 1,2,\ldots$.
Therefore each state in the random walk on an infinite set $S$ is
periodic with period 2.

::: {.example}
*Suppose the states of the system are $\{1,2,3,4 \}$ and
the transition matrix is $$P = \begin{bmatrix}
    0 & 1/2 & 0 & 1/2    \\
    1/4 & 0 & 3/4 & 0   \\
    0 & 1/3 & 0 & 2/3    \\
    1/2 & 0 & 1/2 & 0   \\
\end{bmatrix}.$$ Determine the period of each state.*
:::
**Solution:**
The entire state space is a single class. Note also that
$p^{(2n)}_{ii} > 0$ and $p^{(2n+1)}_{ii} = 0$ for $n = 1,2,\ldots$.
Therefore each state in the state space $S$ is
periodic with period 2.

::: {#exampleLongTerm .example}
*A Markov chain $X_0, X_1, \ldots$ on states $\{1,2\}$
has the following transition matrix $$P = \begin{bmatrix}
    0 & 1   \\
    1 & 0   \\
    %\vdots & \vdots & \vdots  & \vdots \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix},$$*

1.  Find the stationary
    distribution(s) of this Markov chain.*

2.  *Describe the long-term behaviour of the Markov chain. Does the
    distribution of the chain tend to the stationary distribution(s)
    found in 1.*
:::
**Solution:**
1.  To find the stationary distribution
    $\boldsymbol{\pi} = (\pi_{1},  \pi_{2})$, we need to solve
    $$\left(\pi_{1},  \pi_{2}\right) \begin{bmatrix}
        0 & 1   \\
        1 & 0   \\
        %\vdots & \vdots & \vdots  & \vdots \\
        %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
    \end{bmatrix} = \left(\pi_{1},  \pi_{2}\right),$$ and
    $$\pi_{1} +  \pi_{2}  = 1.$$ This gives
    $\boldsymbol{\pi} = (1/2, 1/2).$

2.  There is an equal chance of being in either state. Note that for any
    initial probability distribution
    $\boldsymbol{\mu} =  (\mu_{1},\mu_{2})$ with $\mu_{1}+\mu_{2}=1$, we
    have

    $$\boldsymbol{\mu} \cdot P=\boldsymbol{\mu} \cdot P^{3}=\boldsymbol{\mu} \cdot P^{5}=\ldots
    =(\mu_{2},\mu_{1})$$ and
    $$\boldsymbol{\mu} P^{2}=\boldsymbol{\mu} P^{4}=\boldsymbol{\mu} P^{6}=\ldots = (\mu_{1},\mu_{2}).$$

    The process does not settle down to an equilibrium position. Note
    also that the chain is not aperiodic, i.e. each state is periodic of
    period 2. The process does not conform to stationary in the long
    run.

::: {.example}
*A Markov chain $X_0, X_1, \ldots$ on states $\{1,2\}$
has the following transition matrix $$P = \begin{bmatrix}
    1/3 & 2/3   \\
    2/3 & 1/3   \\
    %\vdots & \vdots & \vdots  & \vdots \\
    %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
\end{bmatrix},$$ Answer the same questions as given in the Example \@ref(exm:exampleLongTerm)*
:::
**Solution:**
1.  The process is finite and irreducible, so a unique stationary
    distribution exists. To find the stationary distribution
    $\boldsymbol{\pi} = (\pi_{1},  \pi_{2})$, we need to solve
    $$\left(\pi_{1},  \pi_{2}\right) \begin{bmatrix}
        1/3 & 2/3   \\
        2/3 & 1/3   \\
        %\vdots & \vdots & \vdots  & \vdots \\
        %p_{d1} & p_{d2} & p_{d3} & \dots  & p_{dn}
    \end{bmatrix} = \left(\pi_{1},  \pi_{2}\right),$$ and
    $$\pi_{1} +  \pi_{2}  = 1.$$ Solving the system of linear equations
    gives $\boldsymbol{\pi} = (1/2, 1/2).$

2.  The chain is aperiodic. Therefore, according to the results from
    Example \@ref(exm:exampleStationary), if follows that
    $$\lim_{n \rightarrow \infty} p_{i1}^{(n)} = 1/2 > 0 , \text{ and }  \lim_{n \rightarrow \infty} p_{i2}^{(n)} = 1/2,$$
    which is independent of $i$. This is contrast to the process given
    in Example [Example 30](#exampleLongTerm){reference-type="ref"
    reference="exampleLongTerm"}, i.e. the process in this example
    reaches the stationary probability distribution in the long run.

##  Main result {#main-result}


The main result in this section can be stated as follows:

::: {.theorem}
**Theorem 3**. *Let $P$ be the transition probability matrix of a
homogeneous discrete-time Markov chain $\{X_n \}_{n \ge 0}$. If the
Markov chain is*

-   *finite,*

-   *irreducible and*

-   *aperiodic,*

*then there is the unique probability distribution
$\boldsymbol{\pi} = (\pi_j)_{j\in S}$ such that
$$\lim_{n \rightarrow \infty} p^{(n)}_{ij} = \pi_j > 0, \text{ for any } j \in S$$
and $$\sum_{j \in S} \pi_j = 1,$$ and this distribution is independent
of the initial state. Such probability distribution $\boldsymbol{\pi}$
is called the **limiting probability distribution**. In addition, the
limiting distribution $\boldsymbol{\pi} = (\pi_j)_{j\in S}$ is the
stationary probability distribution of the Markov chain, i.e. it also
satisfies $\boldsymbol{\pi}  P  =\boldsymbol{\pi}$.*
:::

::: {.example}
*Recall the Markov chain as given in
Example [Example 22](#exampleStationary){reference-type="ref"
reference="exampleStationary"}. The Markov chain is finite, irreducible
and aperiodic. We have also shown that
$$\boldsymbol{\pi} = \left( \frac{b}{a+b},  \frac{a}{a+b} \right)$$ is
the limiting distribution, i.e.
$$\lim_{n \rightarrow \infty} p_{ij}^{(n)} = \pi_j > 0 , \text{ for }j \in \{1,2\},$$
independent of $i$. This limiting distribution is also the unique
stationary distribution of the Markov chain, which can be verified by
$$\left( \frac{b}{a+b},  \frac{a}{a+b} \right) \begin{bmatrix}
    1-a & a   \\
    b & 1-b   \\
\end{bmatrix} =  \left( \frac{b}{a+b},  \frac{a}{a+b} \right),$$ where
$0 < a,b < 1.$*
:::

::: {.example}
*The above result can be applied to the NCD system
because it is finite, irreducible and aperiodic. Indeed, there is the
unique limiting probability distribution
$$\boldsymbol{\pi} = \left(\frac{(p-1)^2}{p^2 - p +1}, \frac{p - p^2}{p^2 - p +1}, \frac{p^2}{p^2 - p +1}\right),$$
which is the stationary distribution of the chain. This gives the
long-term behaviour of the Markov chain, i.e. the probability of finding
the Markov chain in state $j$ is approximately $\pi_j$ independent of
the initial distribution.*

*For example, let $p =   0.8$ in the transition probability matrix $P$.
We compute several powers of $P$ as follows: $$\begin{aligned}
P &= \begin{bmatrix}
0.2 & 0.8 & 0\\
0.2 & 0 & 0.8\\
0 & 0.2 & 0.8\\
\end{bmatrix}
,  &
P^2 &= \begin{bmatrix}
0.2 & 0.16 & 0.64\\
0.04 & 0.32 & 0.64\\
0.04 & 0.16 & 0.8\\
\end{bmatrix}, \\
P^4 &= \begin{bmatrix}
0.072 & 0.1856 & 0.7424\\
0.0464 & 0.2112 & 0.7424\\
0.0464 & 0.1856 & 0.768\\
\end{bmatrix}
,  &
P^8 &= \begin{bmatrix}
0.0482432 & 0.1903514 & 0.7614054\\
0.04758784 & 0.1910067 & 0.7614054\\
0.04758784 & 0.1903514 & 0.7620608\\
\end{bmatrix} \\
P^{16} &= \begin{bmatrix}
0.04761946 & 0.1904761 & 0.7619044\\
0.04761903 & 0.1904765 & 0.7619044\\
0.04761903 & 0.1904761 & 0.7619049\\
\end{bmatrix}
,  &
P^{32} &= \begin{bmatrix}
0.04761905 & 0.1904762 & 0.7619048\\
0.04761905 & 0.1904762 & 0.7619048\\
0.04761905 & 0.1904762 & 0.7619048\\
\end{bmatrix}.\end{aligned}$$*
:::

The limiting probability distribution is
$$\lim_{n\rightarrow \infty} P^n=
 \begin{bmatrix}
0.04761905 & 0.1904762 & 0.7619048\\
0.04761905 & 0.1904762 & 0.7619048\\
0.04761905 & 0.1904762 & 0.7619048\\
\end{bmatrix}.$$

### Applications of Markov chains to NCD systems {#applications-of-markov-chains-to-ncd-systems .unnumbered}

::: {.example}
*A no-claims discount system for motor insurance has
three levels of discount:*

    *Level*     *1*     *2*     *3*
  ------------ ------ ------- -------
   *Discount*   *0%*   *30%*   *50%*

*The rules for moving between these levels are given as follows:*

-   *Following a claim-free year, move to the next higher level, or
    remain at level 3.*

-   *Following a year with one claim, move to the next lower level, or
    remain at level 1.*

-   *Following a year with two or more claims, move to level 1, or
    remain at level 1.*

*A portfolio consists of 10000 policyholders, of which*

-   *5000 policyholders are classified as good drivers. The number of
    claims per year in this group is $\text{Poisson}(0.1)$.*

-   *5000 policyholders are classified as bad drivers. The number of
    claims per year in this group is $\text{Poisson}(0.2)$.*

1.  *Calculate $\Pr[N = 0]$, $\Pr[N = 1]$, and $\Pr[N \ge 2]$ for each
    group.*

2.  *Write down the transition probability matrix of this no-claims
    discount system for each group.*

3.  *Calculate the expected number of policyholders at each level for
    each group once stability has been achieved.*

4.  *Calculate the expected premium income per driver from each group
    once stability has been achieved.*

5.  *Calculate the ratio of the expected premium income per driver from
    the group of good drivers to that from the group of bad drivers once
    stability has been achieved.*

6.  *Comments on the results obtained. Does this NCD system encourage
    good driving?*
:::

